Transition path sampling based calculations of free energies of enzymatic reactions: the case of human methionine adenosyl transferase and plasmodium vivax adenosine deaminase

### 2022-05-03 tex/manuscript.tex
added a whole bunch of images and text

A biased initial reactive trajectory is obtained by using harmonic restraint forces 
of 1500 to break the $O-C$ bond and form the $S-C$ bond for the MAT2A enzyme and 
to break the $N-C$ bond and form the $O-C$ bond for the case of the $pv$ADA enzyme. 
Using the biased trajectory as the starting point and after removing the restraint 
forces, the shooting algorithm was used to generate
new trajectories within which first a random time slice is chosen from the reactive 
trajectory and perturbing the momenta of the system. 
The time step that we used for all our calculations was $1\;fs$ and we propagate the 
dynamics for a total of 250 steps forward and 250 steps backward which amounts to a 
total of $0.5\;ps$. 
A total of 150 and 100 reactive trajectories were generated for the MAT2A and $pv$ADA
enzymes, respectively. 
%------------------------------------------------------------------------------ 
\subsection{Free energy calculations}
%------------------------------------------------------------------------------ 
To obtain the free energy profile along the order parameter, we use the method 
discussed in sec. \ref{ssec:algorithm}. The order paratemer for the MAT2A system 
is defined to be $dis(OC)-dis(SC)$ and that for the $pv$ADA is defined as 
$dis(NC)-dis(OC)$. For the MAT2A system, configurations are sampled within 
the range $[-2.0,2.0]$ and this range is divided into 14 overlapping windows 
($0.09$ {\AA} overlap on either side). 
Configurations are sampled for the $pv$ADA system within the interval $[-3.0, 3.0]$ {\AA}
of the order parameter value. 
This interval is divided into 20 overlapping windows ($0.045$ {\AA} between 
neighboring windows). A reactive trajectory from the TPS ensemble is chosen to be the guide trajectory 
for sampling within each of the windows. To sample in a particular window $w$, a 
frame from this reactive trajectory is chosen such that the value of its 
order parameter falls within $w$. Using the shooting algorithm
the momenta are perturbed and dynamics is run for 10 steps forwards and backwards with a 
$1\;fs$ time step to obtain a trajectory with a time period of $20\;fs$.    
Within each window 3000 TPS trajectories are collected. The 
probability distribution function for the order parameter within window $w$ 
is then calculated as normalized histograms which represent the populations 
of the configurations within the window at equilibrium. 
Then we calculate the free energy profile in a similar fashion to the WHAM 
procedure \cite{Kumar92JComputChem13p1011} which is generally used to obtain the 
potential of mean force from umbrella sampling calculations. 
%------------------------------------------------------------------------------ 
\section{Results and discussion}
%------------------------------------------------------------------------------ 
\subsection{Human Methionine Adenosyltransferase 2A}
The equilibrated structure of the MAT2A enzyme in complex with 
\begin{figure}[ht!]
\includegraphics[scale=0.2]{figures/mat2a-equil.png}
\label{fig:mat2a-equil}
\caption{Equilibriated structure of the MAT2A enzyme in complex with ATP and 
MET along with 2 Mg$^{2+}$ ions.}
\end{figure}
The adenosyl part of the ATP is held together through $\pi-\pi$ interactions 
by Phe250 and Ile117 residues.

\begin{figure}
\includegraphics[scale=1.0]{figures/mat2a-diff167.pdf}
\caption{A reactive trajectory}
\label{fig:mat2a-reactive-traj}
\end{figure}
The order parameters of the system along a typical reactive trajectory is 
shown in Fig. \ref{fig:mat2a-reactive-traj}.

@@

\begin{figure}[ht!]
\includegraphics[scale=0.2]{figures/mat2a-transition-state.png}
%\includegraphics[scale=0.2]{figures/mat2a-active-site2.png}
\end{figure}

From the committor distribution analysis for a reactive trajectory we see that the 
constrained region consisting of just the QM part which is the ATP and MET molecules 
along with two Mg$^{2+}$ ions results in a distribution that is not peaked 
at 0.5 whereas when the residues ... are constrained we obtain a distribution 
that is peaked at 0.5. Therefore we conclude that the reaction coordinate consists
of these residues as well. 
\begin{figure*}[ht!]
\centering
\begin{minipage}[b]{0.45\linewidth}
\includegraphics[width=\textwidth]{figures/comm-60-mat2a-nocons.pdf}
\label{fig:minipage1}
\end{minipage}
\quad
\begin{minipage}[b]{0.45\linewidth}
\includegraphics[width=\textwidth]{figures/comm-60-mat2a.pdf}
\label{fig:minipage2}
\end{minipage}
\caption{Committor distribution analysis}
\end{figure*}

@@

\begin{figure}
\includegraphics[scale=1.0]{figures/mat2a-fenergy.pdf}
\caption{The free energy as a function of the order parameter.}
\end{figure}
From our TPS based calculations we calculated the free energy barrier to be 
$16\;\text{Kcal mol}^{-1}$.
%------------------------------------------------------------------------------ 
\subsection{Adenosine deaminase reaction coordinates and free energy}
%------------------------------------------------------------------------------ 
The equilibrated structure for the ADA enzyme complexed with adenosine, 
OH$^{-}$ anion and Zn$^{2+}$ ion is shown in Fig. \ref{fig:ada-equil}.
The Zn$^{2+}$ cation interacts with His42, His44 and His226 residues
along with the hydroxide ion in the pocket. Glu229 is part of the 
QM region and it acts as a proton donor to the adenosine molecule.

@@

\begin{figure}
\centering
\includegraphics[scale=0.12]{blender-images/ada/ada-trans.png}
\label{fig:ada-trans}
\caption{The transition state structure of the reaction catalyzed by the ADA enzyme}
\end{figure}

\begin{figure}[ht!]
\includegraphics[scale=1.0]{./figures/ada-fenergy.pdf}
\label{fig:ada-fenergy}
\caption{The free energy curve for the ADA reaction}
\end{figure}

>>>>

A biased initial reactive trajectory is obtained by using harmonic restraint forces 
of 1500 to break the $O-C$ bond and form the $S-C$ bond for the MAT2A enzyme and 
to break the $N-C$ bond and form the $O-C$ bond for the case of the $pv$ADA enzyme. 
Using the biased trajectory as the starting point and after removing the restraint 
forces, the shooting algorithm was used to generate
new trajectories within which first a random time slice is chosen from the reactive 
trajectory and perturbing the momenta of the system. 
The time step that we used for all our calculations was $1\;fs$ and we propagate the 
dynamics for a total of 250 steps forward and 250 steps backward which amounts to a 
total of $0.5\;ps$. 
A total of 150 and 100 reactive trajectories were generated for the MAT2A and $pv$ADA
enzymes, respectively. 
%------------------------------------------------------------------------------ 
\subsection{Free energy calculations}
%------------------------------------------------------------------------------ 
To obtain the free energy profile along the order parameter, we use the method 
discussed in sec. \ref{ssec:algorithm}. The order paratemer for the MAT2A system 
is defined to be $dis(OC)-dis(SC)$ and that for the $pv$ADA is defined as 
$dis(NC)-dis(OC)$. For the MAT2A system, configurations are sampled within 
the range $[-2.0,2.0]$ and this range is divided into 14 overlapping windows 
($0.09$ {\AA} overlap on either side). 
Configurations are sampled for the $pv$ADA system within the interval $[-3.0, 3.0]$ {\AA}
of the order parameter value. 
This interval is divided into 20 overlapping windows ($0.045$ {\AA} between 
neighboring windows). A reactive trajectory from the TPS ensemble is chosen to be the guide trajectory 
for sampling within each of the windows. To sample in a particular window $w$, a 
frame from this reactive trajectory is chosen such that the value of its 
order parameter falls within $w$. Using the shooting algorithm
the momenta are perturbed and dynamics is run for 10 steps forwards and backwards with a 
$1\;fs$ time step to obtain a trajectory with a time period of $20\;fs$.    
Within each window 3000 TPS trajectories are collected. The 
probability distribution function for the order parameter within window $w$ 
is then calculated as normalized histograms which represent the populations 
of the configurations within the window at equilibrium. 
Then we calculate the free energy profile in a similar fashion to the WHAM 
procedure \cite{Kumar92JComputChem13p1011} which is generally used to obtain the 
potential of mean force from umbrella sampling calculations. 
%------------------------------------------------------------------------------ 
\section{Results and discussion}
%------------------------------------------------------------------------------ 
\subsection{Human Methionine Adenosyltransferase 2A}
The equilibrated structure of the MAT2A enzyme in complex with 
ATP, MET and two Mg$^{2+}$ ions is shown in Fig. \ref{fig:mat2a-equil}.
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.2]{figures/mat2a-equil.png}
\caption{Equilibriated structure of the MAT2A enzyme in complex with ATP and 
MET along with 2 Mg$^{2+}$ ions.}
\label{fig:mat2a-equil}
\end{figure}
In the binding pocket the adenosyl moiety and the Phe250 and Ile117 residues 
are found to have $\pi-\pi$ interactions stabilizing the molecule. The triphosphate 
moiety forms hydrogen bonds with ... residues.

\begin{figure}
\includegraphics[scale=1.0]{figures/mat2a-diff167.pdf}
\caption{Bond breaking ($dis(OC)$), bond forming ($dis(SC)$) distances for a typical 
reactive trajectory for the reaction catalyzed by the MAT2A enzyme.}
\label{fig:mat2a-reactive-traj}
\end{figure}
The order parameters of the system along a typical reactive trajectory is 
shown in Fig. \ref{fig:mat2a-reactive-traj}.

@@

%\begin{figure}[ht!]
%\includegraphics[scale=0.2]{figures/mat2a-transition-state.png}
%\includegraphics[scale=0.2]{figures/mat2a-active-site2.png}
%\end{figure}

\begin{figure}[ht!]
\includegraphics[scale=0.12]{blender-images/ada/mat2a-trans-labelled.png}
\caption{Transition state structure of the reaction catalyzed by the human MAT2A enzyme.
The S$-$C distance is $2.09$ {\AA} and the O$-$C distance is $2.32$ {\AA} which represents 
an advanced transition state.}
\label{fig:mat2a-trans-struct}
\end{figure}
The transition state structure follows a S$_{\text{N}}2$ 
mechanism with a product like structure, the O$-$C bond is cleaved and 
has a value of $2.32$ {\AA}.
Experimentally the transition state structures of enzymatic reactions
can be elucidated using kinetic isotope effects (KIE). \cite{Schramm99MetEnzym308p301}

From the committor distribution analysis for a reactive trajectory we see that the 
constrained region consisting of just the QM part which is the ATP and MET molecules 
along with two Mg$^{2+}$ ions results in a distribution that is not peaked 
at 0.5 whereas when the residues ... are constrained we obtain a distribution 
that is peaked at 0.5. Therefore we conclude that the reaction coordinate consists
of these residues as well. 
\begin{figure*}[ht!]
\centering
\begin{minipage}[b]{0.45\linewidth}
\includegraphics[width=\textwidth]{figures/comm-60-mat2a-nocons.pdf}
\label{fig:minipage1}
\end{minipage}
\quad
\begin{minipage}[b]{0.45\linewidth}
\includegraphics[width=\textwidth]{figures/comm-60-mat2a.pdf}
\label{fig:minipage2}
\end{minipage}
\caption{Committor distribution analysis for obtaining the reaction coordinate of the 
MAT2A catalyzed reaction. The figure on the left has the QM region constrained and the figure 
on the right has the QM region along with the Gln113, Ser114, Arg249 and Arg264 residues constrained.}
\end{figure*}

@@

\begin{figure}
\includegraphics[scale=1.0]{figures/mat2a-fenergy.pdf}
\caption{The free energy as a function of the order parameter.}
\end{figure}
From our TPS based calculations we calculated the free energy barrier to be 
$16\;\text{Kcal mol}^{-1}$.
%------------------------------------------------------------------------------ 
\subsection{Adenosine deaminase reaction coordinates and free energy}
%------------------------------------------------------------------------------ 
The equilibrated structure for the ADA enzyme complexed with adenosine, 
OH$^{-}$ anion and Zn$^{2+}$ cation is shown in Fig. \ref{fig:ada-equil}.
The Zn$^{2+}$ cation interacts with the His42, His44 and His226 residues
and stabilizes the OH$^{-}$ anion in the pocket.

Glu229 is part of the 
QM region and it acts as a proton donor to the adenosine molecule.

Experimentally KIE studies 
from Schramm and coworkers \cite{Luo07JAmChemSoc129p8008,Ho09Biochemistry48p9618}
predict the transition state structures for the plasmodium falciparum, human and bovine 
ADA enzymes. They reported inhibition studies on the plasmodium vivax ADA by using 
MT coformycin as the inhibitor molecule.

@@

\begin{figure}
\centering
\includegraphics[scale=0.12]{blender-images/ada/new-ada-trans.png}
\label{fig:ada-trans}
\caption{The transition state structure of the reaction catalyzed by the ADA enzyme.
The formation of the Meissenheimer complex is the rate limiting step and the N-C distance is 
1.60 {\AA} and the O-C distance is 1.48 {\AA}.}
\end{figure}

\begin{figure}[ht!]
\includegraphics[scale=1.0]{./figures/ada-fenergy.pdf}
\label{fig:ada-fenergy}
\caption{The free energy in $\text{Kcal}\;\text{mol}^{-1}$ as a function of the reaction 
parameter for the ADA catalyzed reaction. The Meissenheimer complex where the OH$^{-}$ anion
is bonded to the C atom and the leaving NH$_2$ moeity is also still bonded.}
\end{figure}
